Less Funding, More Lecturers, and Fewer Professors: \\ \vspace{0.1cm} \Large{Stagnating State Funding for Higher Education and its Effect on Faculty at Public Universities}

### 2023-12-19 text/paper.tex
adjusted writing
Adjustments.
Wiritng updates, new title.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title page
% Author
\author{Senan Hogan-Hennessy\thanks{
        For helpful comments I thank
        Levon Barseghyan,
        Francine Blau,
        Ryan Dycus,
        Ronald Ehrenberg,
        Michael Lovenheim,
        Jake Meyer,
        Douglas Miller,
        and Evan Riehl,
        as well as seminar participants at Cornell University (Autumn 2022) for helpful discussion.} \\
    \vspace{0.1cm}
    Economics Department, Cornell University
}
% Title
\title{Stagnating State Funding for Higher Education and its Effect on Faculty at US Universities
    }
\date{This version: \today\thanks{
    This project's Github repository, with materials and anonymised data for replication, is available at 
    \url{https://github.com/shoganhennessy/state-funding-faculty}.
    Any comments or suggestions may be sent to me at \href{mailto:seh325@cornell.edu}{\nolinkurl{seh325@cornell.edu}}, or raised as an issue on the Github project.
    } \\ \vspace{0.1cm}
    \href{https://github.com/shoganhennessy/state-funding-faculty/blob/main/state-funding-faculty-2023.pdf}{Newest version available here.}
}
\rhead{\today}
\lhead{Stagnating State Funding for Higher Education and its Effect on Faculty.}
% Begin
\begin{document}
\maketitle
\thispagestyle{empty}
% Abstract
\begin{abstract}
    \noindent
    \input{sections/00-abstract.tex}
\end{abstract}

>>>>

\def\sectionautorefname~#1\null{Section~#1\null}                       % Fix autoref for sections

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title page
% Author
\author{Senan Hogan-Hennessy\thanks{
        For helpful comments I thank
        Levon Barseghyan,
        Francine Blau,
        Ryan Dycus,
        Ronald Ehrenberg,
        Michael Lovenheim,
        Jake Meyer,
        Douglas Miller,
        and Evan Riehl,
        as well as seminar participants at Cornell University for helpful discussion.} \\
    \vspace{0.1cm}
    Economics Department, Cornell University
}
% Title
\title{Less Funding, More Lecturers, and Fewer Professors: \\ \vspace{0.1cm}
    \Large{Stagnating State Funding for Higher Education and its Effect on Faculty at Public Universities}}
\date{This version: \today\thanks{
    This project's Github repository, with materials and anonymised data for replication, is available at 
    \url{https://github.com/shoganhennessy/state-funding-faculty}.
    Any comments or suggestions may be sent to me at \href{mailto:seh325@cornell.edu}{\nolinkurl{seh325@cornell.edu}}, or raised as an issue on the Github project.
    } \\ \vspace{0.1cm}
    \href{https://github.com/shoganhennessy/state-funding-faculty/blob/main/state-funding-faculty-2023.pdf}{Newest version available here.}
}
\rhead{\today}
\lhead{Less Funding, More Lecturers, and Fewer Professors.}
% Begin
\begin{document}
\maketitle
\thispagestyle{empty}
% Abstract
\begin{abstract}
    \noindent
    \input{sections/00-abstract.tex}
\end{abstract}

### 2023-12-19 text/sections/00-abstract.tex
Wiritng updates, new title.

US public universities have systematically substituted away from tenure-track/tenured professors towards lecturers over the last thirty years, while states have systematically cut public funding for higher education.
This paper shows that falling state funding caused changes in faculty composition, using a shift-share approach to instrument for state funding cuts.
Universities employ $4.4$\% more lecturers per student following a $10$\% funding cut, and correspondingly rely less on tenure-track faculty by employing $1.4$\% fewer assistant professors and $1.2$\% fewer full professors per student.
In a secondary analysis, using data on all faculty at Illinois public universities, incumbent professors' salaries, promotion rate, and quit rate are not affected by funding cuts, implying that composition change arose by limiting hiring of new tenure-track/tenured professors.
These results show institutional effects of stagnating public funding for education, exhibiting one possible mechanism for deteriorating student outcomes over this same time period.

>>>>

Public universities employ more lecturers than ever, while state funding for higher education has stagnated for over thirty years.
This paper shows falling state funding caused a substitution away from professors towards lecturers at US public universities, using a shift-share approach to instrument for state funding cuts relative to enrolment.
Universities employ $4.4$\% more lecturers per student following a $10$\% funding cut, relying less on professors by employing $1.4$\% fewer assistant professors and $1.2$\% fewer full professors per student.
Incumbent professors' salaries, promotion rate, and quit rate at Illinois universities are not affected by funding cuts, so that the substitution arose by limiting hiring of new tenure-track/tenured professors.
Stagnating public funding affects public universities as institutions, and likely contributed to deteriorating student outcomes at public universities since the 1990s.

### 2023-12-19 text/sections/01-introduction.tex
Wiritng updates, new title.

Public universities have experienced a stagnation in state funding over the last three decades, despite a dramatic increase in enrolments.
At the same time, public unviersities have increasingly subsituted away from tenure-track faculty towards lecturers, who often work on contingenet and lower-salaried contracts.
Previous research has examined the effects of the decline in state funding on student outcomes \citep{NBERw23736,NBERw27885}, but there is little evidence on whether this funding decline has impacted faculty.
In this paper, I show there is a causal relation between the trends of declining state funding for higher education and increasing reliance on lecturers.
In a secondary analysis of all public university faculty in the state of Illinois, I show  shows that incumbent faculty were relatively unaffected, in terms of wages, promotion and quit rates.
This implies that changes in faculty composition arose by disrupting public universities' ability to hire or replace their tenure-track/tenured faculty. 
Private universities were not exposed to similar financial constraints during this same time period, and do not exhibit the substitution away from tenure-track/tenured professors.
Thus stagnating state funding for public universities has implications for the wider landscape of US higher education and research.

State funding for public universities has dramatically declined over recent decades, and the composition of faculty at these universities changed drastically over the same time frame.
This fall is driven by a stagnation in the absolute level of state funding for higher education in most states, combined with a large increase of 46\% in student enrolment at public universities over the same time period.
Similarly, the number of lecturers per student over doubled, while the number of tenure-track/tenured professors per students fell by $-23$\% from 1990 to 2021.
%While US universities are widely considered the highest performing in the world, there are consequential differences between its universities that operate in the private sector and those established by state governments.
Public universities rely on their state governments for funding, and receive over a quarter of their total revenue from state governments.
An average public university received around \$11,600 of state funding per student in 1990, and only \$8,300 in 2021.

It is a priori unclear how faculty may be affected, as there are various margins that universities may to respond to funding cuts.
University administrations may directly cut salaries or lay-off incumbent staff; in 2015, the University of Illinois furloughed faculty around a budget crisis \citep{furlough2010}.
Similarly, faculty hiring committees are often cancelled or disrupted in response to funding cuts, meaning there are fewer faculty entering a department to replace those leaving.
On the other hand, some faculty positions may grow at universities in response to funding cuts.
Public universities do not restrict enrolment in response to funding cuts, so someone must teach the next year of students.
Lecturers can fill in teaching duties, at a lower cost for the universities than tenure-track professors, with less employment protections (e.g., no tenure).
However, over the same time period, higher education has undergone dramatic changes in the rate of selectivity thanks to greater competition \citep{hoxby2009changing}.
This selectivity pressure may temper universities from relying on lecturers, and instead lead them to fill in the funding gap by raising revenue from other sources, such as private funds or raising tuition.

I use two different analyses to explore the causal effects of state funding on faculty composition and salaries.
My first analysis uses data from the \cite{ipeds} to study the impact of changes in state funding for all US public universities 1990--2017.
My second analysis uses panel data for every faculty member at Illinois public universities for 2010--2021 \citep{ibhed}, to measure how incumbent faculty (i.e., those already employed at the university) are affected by changes in state funding.
These data allow me to answer the following research question: if a US public university receives an extra \$1,000 for each of their students (or a 10\% increase), do they change the composition of faculty they employ?
If so, which positions do universities substitute away from, or towards, and do these funding cuts affect the faculty members themselves?

In both analyses, I employ a shift-share instrument to identify changes in state funding for higher education, which exploits how universities differ in their financial reliance on state funding, combined with yearly shocks to the amount an entire state funds higher education --- following \cite{NBERw23736,NBERw27885}.
State governments decide how to fund higher education by a complicated process which can be influenced by local economic conditions, changing state priorities, or even lobbying from the state universities themselves.
As such, state funding is not randomly assigned to universities and so it is necessary to employ an identification strategy, such as the shift-share instrument, to account for this.
The shift-share instrument interacts how much a university relies on state funding as a percent of revenues in a base period with the entire state's total funding per higher education student for each year, exploiting changes in funding and how much each university historically relied on state funding for higher education.
Additionally, I estimate effect of state funding on faculty multiple years after the initial funding cuts, using the local projections method thanks to the presence of time-series confounding between the funding shock and later years' level of state funding.\footnote{
    To the best of my knowledge, this is also the first paper to use the local projections method for a shift-share instrument, and to acknowledge the time-series confounding for shift-share IV models when the outcome is non-contemporaneous.
}

In my national-level analysis, I find that state funding cuts cause public universities to substitute away from tenured and tenure-track professors towards contingent lecturers.
A funding cut of \$1,000 per student leads to an average university employing 6 more lecturers;
in percentage terms, a funding cut of 10\% per student leads to a fall of 1.4\% in the number of assistant professors per student at a university, a fall of 1.2\% for tenured professors, and an increase in 4.4\% the number of lecturers per student.
Local projection estimates show that these effects linger for three to four years after the initial funding shock, showing that the effect is not isolated to the year of the funding shock. 
Over the same time period, state funding per student fell by around 35\%, count of (tenure-tack/tenured) professors per student fell by 9\%, and lecturers per student increased by 99\%, so that these results show that falls in state funding explain around 53\% of the fall in professors per student
% i.e. -(35 * 0.137) / -9
and 15.5\% of the rise in lecturers per student.
% ie., (35 * 0.437) / 99

In my analysis of all Illinois faculty, I find that incumbent professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system.
Yet the hiring rate for new professors at public universities was negatively impacted by the state funding cuts.
This implies that faculty composition change arose by limiting the hiring of new professors; public universities increasingly hired contingent lecturers, and increasingly did not replace their retiring (or leaving) professors.

Mine is the first paper to provide causal evidence on the impact of state funding on faculty composition and wages.
The closest related research on higher education funding has examined how university spending affected graduation rates and levels of student debt \citep{NBERw23736,NBERw27885}, and university finances \citep{miller2022making,bound2019public,brown2014endowment}.
For faculty outcomes, previous applied theory work models university decision-making and faculty hiring (see e.g., \citealt{abe2015implications,johnson2009jep,NBERc13879}).
In addition, several empirical papers measure how universities were effected by endowment shocks \citep{brown2014endowment}, and trends in the academic job market following the 2008 recession \citep{turner2014impact}.
These papers measure changes in state funding and university revenues, but do not measure effects on faculty composition or salaries, or other outcomes related to university instruction.

Another contribution of my paper is to provide evidence on possible mechanisms for the previous research connecting state funding cuts for higher education and worsening student outcomes.
Substitution towards lecturers is likely a cost-cutting measure for funding constrained universities, though relying on these faculty --- who are often over-worked and not granted long-term employment protections --- may lead to worse student outcomes \citep{ehrenberg2005tenured,zhu2021limited,jaeger2011examining}.
Lastly, my results provide evidence that the long term trends in higher education funding are causally related, contributing to the literature in the long run trends in faculty outcomes \citep{ehrenberg2003studying} and trends in US higher education and funding \citep{hoxby2009changing}.

This paper proceeds as follows.
\autoref{sec:data} describes the data for university finances and faculty in Illinois, and trends in public university funding for the last three decades.
\autoref{sec:conceptual} gives the conceptual framework for how state funding may affect faculty.
\autoref{sec:empirics} draws the empirical framework for isolating the causal effects of state funding on faculty composition and individual faculty, and \autoref{sec:results} presents the empirical results.
\autoref{sec:discussion} discusses the context and implications for the findings.
\autoref{sec:conclusion} concludes.

>>>>

States fund their public universities with roughly the same amount today as they did in the 1990s, despite their enrolment rising by over 50\% on average.
At the same time, public universities have systematically substituted away from professors towards lecturers, who teach often work on contingent and lower-salaried contracts.
Previous research shows declining in state funding led to worsening student outcomes \citep{NBERw23736,NBERw27885}, but there is little evidence on whether this funding decline has impacted faculty.
In this paper, I show that falling state funding caused a substitution away from professors toward lecturers at US public universities.

Per student state funding for higher education has stagnated the last thirty years.
The average public university received around \$11,600 of state funding per student in 1990 and \$8,300 in 2021, while the number of lecturers per student increased by 109\% and professors per student decreased by $23$\% over 1990--2021.
I use a shift-share instrumental variables approach to estimate how faculty are affected by these funding cuts.
Universities employ $4.4$\% more lecturers per student following a $10$\% funding cut, and correspondingly rely less on tenure-track faculty by employing $1.4$\% fewer assistant professors and $1.2$\% fewer full professors per student.
In a secondary analysis of all public university faculty in the state of Illinois, I show that incumbent faculty were relatively unaffected, in terms of wages, promotion and quit rates.
This implies that the faculty substitution arose by disrupting hiring of new tenure-track/tenured professors.
% Private universities were not exposed to similar financial constraints during this same time period, and do not exhibit the substitution away from tenure-track/tenured professors.
This paper shows the institutional effects of stagnating state funding for higher education, and offers one possible mechanism for how funding cuts affected student outcomes.

I use two different analyses to explore the causal effects of state funding on faculty composition and salaries.
My first analysis uses data from the \cite{ipeds} to study the impact of changes in state funding for all US public universities 1990--2021.
My second analysis uses panel data for every faculty member at Illinois public universities for 2010--2021 \citep{ibhed}, to measure how incumbent faculty (i.e., those already employed at the university) are affected by changes in state funding.
These data allow me to answer the following research question: if a US public university receives an extra \$1,000 per student (or a 10\% increase), do they change how many faculty they employ?
If so, which positions do universities substitute away from, or towards, and do these funding cuts affect the faculty themselves?

It is a priori unclear how faculty may be affected, as there are various margins that universities may to respond to funding cuts.
For example, university administrations may directly cut salaries or lay-off incumbent staff.
This happened in 2015 when the state of Illinois shut-down without setting a state budget, and the University of Illinois furloughed faculty thanks to the funding shortfall \citep{furlough2010}.
Similarly, faculty hiring committees are often cancelled or disrupted in response to funding cuts, meaning there are fewer faculty entering a department to replace those leaving.
On the other hand, some faculty positions may grow at universities in response to funding cuts.
Public universities do not restrict enrolment in response to funding cuts, so someone must teach the next year of students.
Lecturers can fill in teaching duties, at a lower cost for the universities than tenure-track professors, with less employment protections (e.g., no tenure).
But it is not certain that universities would hire more lecturers in the face of budget cuts; they may not hire any more instructional staff thanks to lack of resources.
Similarly, the universities may be limited from hiring more lecturers by competitive pressures from peer institutions \citep{hoxby2009changing,urquiola2020markets}, and instead focus on filling the funding gap by raising revenue from other sources, such as private funds or raising tuition.

In both analyses, I employ a shift-share instrument to identify changes in state funding for higher education.
State governments decide how to fund higher education by a complicated process which can be influenced by local economic conditions, changing state priorities, or informal lobbying from universities.
As such, state funding is not randomly assigned to universities and so it is necessary to employ an identification strategy to account for this.
The shift-share instrument identifies exogenous changes in state funding by exploiting differences in how much each university relied on state funding, interacted with state-wide changes in higher education funding --- following \cite{NBERw23736,NBERw27885}.
The approach relies on the insight that yearly state government budgets are decided without targetting individual universities for funding cuts, and exploits the fact that more reliant universities will be more affected by state-wide changes in higher education funding.
Additionally, I estimate effect of state funding on faculty multiple years after the initial funding cuts, using the local projections method thanks to time-series confounding between the funding shock and later years' level of state funding.\footnote{
    To the best of my knowledge, this is also the first paper to use the local projections method for a shift-share instrument, and to acknowledge the time-series confounding for shift-share IV models when the outcome is in future time periods.
}

In my national-level analysis (IPEDS), I find that state funding cuts cause public universities to substitute away from tenured and tenure-track professors towards lecturers.
A funding cut of \$1,000 per student leads to an average university employing 6 more lecturers;
in percentage terms, a funding cut of 10\% per student leads to a fall of 1.4\% in the number of assistant professors per student at a university, a fall of 1.2\% for tenured professors, and an increase in 4.4\% the number of lecturers per student.
Local projection estimates show that these effects linger for many years after the initial funding shock, showing that the effect is not isolated to the year of the funding cut. 
Over the same time period, state funding per student fell by around 35\%, professors per student fell by 9\%, and lecturers per student increased by 99\%.
These results show that falls in state funding explain around 53\% of the fall in professors per student,
% i.e. -(35 * 0.137) / -9
and 15.5\% of the rise in lecturers per student.
% ie., (35 * 0.437) / 99
In my analysis of all Illinois faculty (IBHED), I find that incumbent professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system.
Yet the hiring rate for new professors at public universities was negatively impacted by the state funding cuts.
This implies that faculty composition change arose by limiting the hiring of new professors; public universities increasingly hired contingent lecturers, and increasingly did not replace their retiring (or otherwise leaving) professors.
Suggestive evidence using data on cumulative hiring counts supports this hypothesis.

Mine is the first paper to provide causal evidence on the impact of state funding on faculty composition and wages.
The closest related research on higher education funding examined how university spending affected graduation rates and levels of student debt \citep{NBERw23736,NBERw27885}, and university finances \citep{miller2022making,bound2019public,brown2014endowment}.
For faculty outcomes, applied theory work models university decision-making and faculty hiring (see e.g., \citealt{abe2015implications,johnson2009jep,NBERc13879}).
In addition, empirical papers measure how universities responded to endowment shocks \citep{brown2014endowment}, and trends in the academic job market following the 2008 recession \citep{turner2014impact}.
These papers measure changes in state funding and university revenues, but do not measure effects on faculty composition or salaries, or other outcomes related to university instruction.

Another contribution of my paper is to provide evidence on possible mechanisms connecting state funding cuts for higher education and worsening student outcomes.
Substitution towards lecturers is likely a cost-cutting measure for funding constrained universities, though relying on these faculty --- who are often over-worked and not granted long-term employment protections --- may lead to worse student outcomes \citep{ehrenberg2005tenured,zhu2021limited,jaeger2011examining}.
Lastly, my results provide evidence that the long term trends in higher education funding are causally related, contributing to the literature on faculty outcomes \citep{ehrenberg2003studying} and trends in US higher education and funding \citep{hoxby2009changing,ehrenberg2012american}.

This paper proceeds as follows.
\autoref{sec:data} describes the data for university finances and faculty in Illinois, and trends in public university funding for the last three decades.
\autoref{sec:conceptual} gives the conceptual framework for how state funding may affect faculty.
\autoref{sec:empirics} draws the empirical framework for isolating the causal effects of state funding on faculty composition and individual faculty, with \autoref{sec:results} presenting the empirical results.
\autoref{sec:discussion} discusses the context and implications for the findings.
\autoref{sec:conclusion} concludes.

### 2023-12-19 text/sections/02-data.tex
Wiritng updates, new title.

IPEDS provides information at the university level, but only provides aggregated information within a university.\footnote{
    IPEDS additionally gives a mean salary measure for faculty by rank (inconsistently) for each university.
    Unfortunately, this measure of professors' salaries is particularly crude in measuring on professors' salaries and other individual-outcomes; summary statistics on IPEDS data do not agree with trends in average professor salary over the sample time period compared to summary statistics provided in \citealt{aau2021survey}.
    Similarly, IPEDS reports of the average salary paid to professors at private institutions also disagree with other sources (including those calculated for in Illinois in the IBHED data later in this section) so are not considered in this analysis.
}
To investigate the distribution, I integrate individual-level data for every public university professor in the state of Illinois between the years 2010-2021.
IBHED hosts the information;
Public Act 96-0266 (effective 1 January 2010) requires that each university report base salary and benefits all administrators, faculty members, and instructors employed by the college or university.\footnote{
    The universities included are all campuses of the nine Illinois public universities: Chicago State University, Eastern Illinois University, Governors State University, Illinois State University, Northeastern Illinois University, Northern Illinois University, Southern Illinois University  (all five campuses), University of Illinois (all four campuses), Western Illinois University.
}
These publicly available data provides the basis to build a panel of Illinois public university professors 2010-2021; I define a professor as an individual by their first plus last name and university pairing, and link this database to IPEDS regarding finances for their employing institution.
The Illinois sample represents 16,932 professors in the year 2010 and 15,352 in the year 2021, with summary statistics presented in \autoref{tab:illinois-summary}.
Additionally, further analysis using the first year of a professor's employment focuses on the subset of professors with observed year of hiring between the years 2011 and 2021, representing 1,778 professors in 2011, and 9,099 in 2021.

@@

The institutional environment for how states appropriate funding allows for wide variation in state funding for higher education, both between states and across years in each state.
A state government plans an annual budget a couple of years ahead of the fiscal year, and the legislature chooses to approve or reject a budget request put forth by the governor's office.\footnote{
    \cite{NBERw23736} present a full discussion of the decision-making process for state appropriations, drawing on administrative records originally analysed by \cite{parmley2009state}.
}
US state governments, by majority, are legally obligated to run a balanced-budget, so that yearly variation in tax revenues (caused by changing economic conditions or otherwise) necessarily affect state expenditures.
This process leads to yearly variation in state funding not seen in other revenues sources (such as federal government appropriations), since US states differ in their support for higher education, and are subject to fiscal constraints.
State funding for higher education are a particularly attractive area of state spending to absorb negative shocks to state finances \citep{delaney2011state}.\footnote{
    \cite{delaney2011state} fully describe the financial environment of state expenditures, and what makes spending on higher education an attractive area for state governments to expand funding during years of higher tax revenues, and retract funding in leaner years.
    An analysis of state expenditures for the years 1980-2004 (overlapping with the sample for this analysis) provides solid evidence for these trends, and \autoref{fig:funding} observes these same trends.
}
Additionally, there is yearly variation in the number of higher education students in each state \citep{turner2014impact}, so that per-student state funding varies on multiple dimensions.

Public university revenues on average stagnated across there years 1990--2021.
\autoref{fig:funding} show the trends in revenues for the mean public university for the years 1990--2021.
We see a rise in total revenues received by public universities (from all sources), and a notable increase in mean tuition revenues from \$48 million per year-university to \$150 million per year-university.
At the same time, total state funding stagnated at around \$100 million per year-university for 1990--2008, falling around 2008 and have not recovered ever since.
While public universities experienced a stagnation in state support, private universities were not exposed to the same constraints, receiving \$37,000 per student in 1990 and \$49,000 in 2021, experiencing no corresponding decline in any specific component.

@@

There are large differences in the average number of professors per student between the private and public sector.
\autoref{fig:fte-perprof} shows that private universities start with  a higher baseline of around 4.5 professors per 100 students, and exhibited yearly variation of less than 0.5 professor per hundred students over the thirty years.
Public universities start with 3.9 professors per hundred students, and this number falls to around 3., with the largest fall in the 2008-2011 time period.
We see a similar difference in baseline, and fall for the years 2008-2011 for assistant professors.
Private and public universities have similar numbers of associated and full professors before the year 2000, yet this number has fallen by over 20\% in the next 20 years only for public universities: in 2021 the mean public university has 6 fewer full professors per hundred students than the mean private university.
Over the same time period, we see the rise in use of non-tenure track instructor positions (referred to as lecturers from here on), who were employed at similar rates in both sectors in 1990, yet have been utilised by public universities at a higher rate since.

@@

2010--2021 saw not only a decline, but also large yearly funding variation, particularly around 2016.
In the calendar year 2015, partisan disagreements between the democratic legislature and republican governor led to the 2016 fiscal year starting with no state budget.
State agencies, and higher education institutions, employed accounting techniques to continue operating without any resources provided by the state government.\footnote{
    Fiscal year 2016 refers to June 2015 to June 2016, so is the same as the academic year definition.
}
While public universities were able to stay open, there were drastic revenue and spending cuts in response to the budget impasse, as it continued through fiscal year 2017, and ended with a new budget restoring funding to state institutions for 2018.
This variation in public university revenues via the state funding channel, stemming entirely from political disagreements --- not from state decisions regarding higher education and its finances \citep{young2020squandered} --- mean that Illinois public universities exhibit sizable changes in their state funding over 2010-2021, of similar order to those for the rest of the country over 1990--2021.

>>>>

IPEDS provides information at the university level, but only provides aggregated information within a university.\footnote{
    IPEDS additionally gives a mean salary measure for faculty by rank (inconsistently) for each university.
    Unfortunately, this measure of professors' salaries is particularly crude in measuring on professors' salaries and other individual-outcomes; summary statistics on IPEDS data do not agree with trends in average professor salary over the sample time period compared to summary statistics provided in \citealt{aau2021survey}.
    Similarly, IPEDS reports of the average salary paid to professors at private institutions also disagree with other sources (including those calculated for Illinois in IBHED data later in this section) so are not considered in this analysis.
}
To investigate the distribution, I integrate individual-level data for every public university professor in the state of Illinois between the years 2010-2021.
IBHED hosts the information;
Public Act 96-0266 (effective 1 January 2010) requires that each university report base salary and benefits all administrators, faculty members, and instructors employed by the college or university.\footnote{
    The universities included are all campuses of the nine Illinois public universities: Chicago State University, Eastern Illinois University, Governors State University, Illinois State University, Northeastern Illinois University, Northern Illinois University, Southern Illinois University  (all five campuses), University of Illinois (all four campuses), Western Illinois University.
}
These publicly available data provides the basis to build a panel of Illinois public university professors 2010-2021; I define a professor as an individual by their first plus last name and university pairing, and link this database to IPEDS regarding finances for their employing institution.
The Illinois sample represents 16,932 professors in the year 2010 and 15,352 in the year 2021, with summary statistics presented in \autoref{tab:illinois-summary}.
Additionally, further analysis using the first year of a professor's employment focuses on the subset of professors with observed year of hiring between the years 2011 and 2021, representing 1,778 professors in 2011, and 9,099 in 2021.

@@

States vary largely in how much they fund their public university systems, thanks in part to the state-wide budgeting process.
Planning for an annual budget begins two years ahead of the fiscal year, and the legislature votes to approve or reject the governor office's budget request a number of months before the fiscal year begins.\footnote{
    \cite{NBERw23736} present a full discussion of the decision-making process for state appropriations, drawing on administrative records originally analysed by \cite{parmley2009state}.
}
US state governments, by majority, are legally obligated to run a balanced-budget, so that yearly variation in tax revenues (caused by changing economic conditions or otherwise) necessarily affect state expenditures.
Public universities have relatively lower lobbying power than other funding beneificaries, so often bear the brunt of state government funding cuts \citep{delaney2011state}.\footnote{
    \cite{delaney2011state} fully describe the financial environment of state expenditures, and what makes spending on higher education an attractive area for state governments to expand funding during years of higher tax revenues, and retract funding in leaner years.
    An analysis of state expenditures for the years 1980-2004 (overlapping with the sample for this analysis) provides solid evidence for these trends, and \autoref{fig:funding} observes these same trends.
}
These features lead to yearly variation in state funding not seen in other revenues sources (such as federal funding), since US states differ in their support for higher education, and are subject to fiscal constraints.
Additionally, the number of higher education students in each state varies thanks to the size of each birth cohort.
For example, the birth cohort of 1971 was larger than for neighbouring years, leading to more student demand for limited public university places for students turning 18 in 1989 \citep{bound2007cohort}.
This leads to variation in the amount of funding per student across years, showing that per student state funding varies on multiple dimensions.

Public university revenues on average stagnated over 1990--2021.
\autoref{fig:funding} show the trends in revenues for the mean public university for the years 1990--2021.
We see a rise in total revenues received by public universities (from all sources), and a notable increase in mean tuition revenues from \$48 million per year-university to \$150 million per year-university.
At the same time, total state funding stagnated at around \$100 million per year-university for 1990--2008, falling around 2008 and have not recovered ever since.
While public universities experienced a stagnation in state support, private universities were not exposed to the same constraints, receiving \$37,000 per student in 1990 and \$49,000 in 2021, experiencing no corresponding decline in any specific component.

@@

There are large differences in the average number of professors per student between the private and public sector.
\autoref{fig:fte-perprof} shows that private universities start with a higher baseline of around 4.5 professors per hundred students, and exhibited yearly variation of less than 0.5 professor per hundred students over the thirty years.
Public universities start with 3.9 professors per hundred students, and this number falls to around 3., with the largest fall in the 2008-2011 time period.
We see a similar difference in baseline, and fall for the years 2008-2011 for assistant professors.
Private and public universities have similar numbers of associated and full professors before the year 2000, yet this number has fallen by over 20\% in the next 20 years only for public universities: in 2021 the mean public university has 6 fewer full professors per hundred students than the mean private university.
Over the same time period, we see the rise in use of non-tenure track instructor positions (referred to as lecturers from here on), who were employed at similar rates in both sectors in 1990, yet have been utilised by public universities at a higher rate since.

@@

There was not only a stagnation in state funding in this time period, but also large annual rises and falls, particularly around 2016.
In the calendar year 2015, partisan disagreements between the democratic legislature and republican governor led to the 2016 fiscal year starting with no state budget.
State agencies, and higher education institutions, employed accounting techniques to continue operating without any resources provided by the state government.\footnote{
    Fiscal year 2016 refers to June 2015 to June 2016, so is the same as the academic year definition.
}
While most public universities were able to stay open, there were drastic revenue and spending cuts in response to the budget impasse, as it continued through fiscal year 2017, and ended with a new budget restoring funding to state institutions for 2018.
This means that Illinois public universities exhibit sizable changes in their state funding over 2010--2021, of similar order to those for the rest of the country over 1990--2021.
Additionally, the 2016 episode stemmed entirely from political disagreements, and not from state decisions regarding higher education and its finances \citep{young2020squandered}, exhibiting how state-level changes in funding affect public universities thanks to unrelated issues.
